---
title: "Hybrid imputation"
author: "Gerko Vink and Stef van Buuren"
date: "Recent advancements in iterative imputation"
output:
  ioslides_presentation:
    logo: mice.png
    smaller: yes
    toc: yes
  beamer_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(kableExtra)
library(mice)
library(dplyr)
library(miceadds)
```

<style>

.notepaper {
  position: relative;
  margin: 30px auto;
  padding: 29px 20px 20px 45px;
  width: 680px;
  line-height: 30px;
  color: #6a5f49;
  text-shadow: 0 1px 1px ;
  background-color: #f2f6c1;
  background-image: -webkit-radial-gradient(center, cover, rgba(255, 255, 255, 0.7) 0%, rgba(255, 255, 255, 0.1) 90%), -webkit-repeating-linear-gradient(top, transparent 0%, transparent 29px, rgba(239, 207, 173, 0.7) 29px, rgba(239, 207, 173, 0.7) 30px);
  background-image: -moz-radial-gradient(center, cover, rgba(255, 255, 255, 0.7) 0%, rgba(255, 255, 255, 0.1) 90%), -moz-repeating-linear-gradient(top, transparent 0%, transparent 29px, rgba(239, 207, 173, 0.7) 29px, rgba(239, 207, 173, 0.7) 30px);
  background-image: -o-radial-gradient(center, cover, rgba(255, 255, 255, 0.7) 0%, rgba(255, 255, 255, 0.1) 90%), -o-repeating-linear-gradient(top, transparent 0%, transparent 29px, rgba(239, 207, 173, 0.7) 29px, rgba(239, 207, 173, 0.7) 30px);
  border: 1px solid #c3baaa;
  border-color: rgba(195, 186, 170, 0.9);
  -webkit-box-sizing: border-box;
  -moz-box-sizing: border-box;
  box-sizing: border-box;
  -webkit-box-shadow: inset 0 1px rgba(255, 255, 255, 0.5), inset 0 0 5px #d8e071, 0 0 1px rgba(0, 0, 0, 0.1), 0 2px rgba(0, 0, 0, 0.02);
  box-shadow: inset 0 1px rgba(255, 255, 255, 0.5), inset 0 0 5px #d8e071, 0 0 1px rgba(0, 0, 0, 0.1), 0 2px rgba(0, 0, 0, 0.02);
}

.notepaper:before, .notepaper:after {
  content: '';
  position: absolute;
  top: 0;
  bottom: 0;
}

.notepaper:before {
  left: 28px;
  width: 2px;
  border: solid #efcfad;
  border-color: rgba(239, 207, 173, 0.9);
  border-width: 0 1px;
}

.notepaper:after {
  z-index: -1;
  left: 0;
  right: 0;
  background: rgba(242, 246, 193, 0.9);
  border: 1px solid rgba(170, 157, 134, 0.7);
  -webkit-transform: rotate(2deg);
  -moz-transform: rotate(2deg);
  -ms-transform: rotate(2deg);
  -o-transform: rotate(2deg);
  transform: rotate(2deg);
}

.quote {
  font-family: Georgia, serif;
  font-size: 14px;
}

.curly-quotes:before, .curly-quotes:after {
  display: inline-block;
  vertical-align: top;
  height: 30px;
  line-height: 48px;
  font-size: 50px;
  opacity: .2;
}

.curly-quotes:before {
  content: '\201C';
  margin-right: 4px;
  margin-left: -8px;
}

.curly-quotes:after {
  content: '\201D';
  margin-left: 4px;
  margin-right: -8px;
}

.quote-by {
  display: block;
  padding-right: 10px;
  text-align: right;
  font-size: 13px;
  font-style: italic;
  color: #84775c;
}

.lt-ie8 .notepaper {
  padding: 15px 25px;
}

</style>

## This presentation | has a website: 

<center>
[www.gerkovink.com/London2019/](https://www.gerkovink.com/London2019/)
</center>
<br><br><br>
You can find all related materials, links and references there. 

# A short overview 

## Imputation
For those of you who are unfamiliar with imputation:

<div class="notepaper">
  <figure class="quote">
    <blockquote class="curly-quotes" cite="https://www.youtube.com/watch?v=qYLrc9hy0t0">
    <font color="black">
    With imputation, some estimation procedure is used to impute (fill in) each missing datum, resulting in a completed dataset that can be analyzed as if the data were completely observed.
    </font>
    </blockquote>
  </figure>
</div>

We can do this once (single imputation) or multiple times (multiple imputation). 

- **With MI, each missing datum is imputed $m \geq 2$ times, resulting in $m$ completed datasets.**

Multiple imputation (Rubin, 1987) has some benefits over single imputation:

- it accounts for missing data uncertainty
- it accounts for parameter uncertainty
- additional adjustments to obtain valid inference are not necessary

## How to go about this?
Once we start the process of multiple imputation, we need a scheme to solve for multivariate missingness

Some notation:

- Let $Y$ be an incomplete column in the data

  - $Y_\mathrm{mis}$ denoting the unobserved part
  - $Y_\mathrm{obs}$ denotes the observed part

- Let $X$ be a set of completely observed covariates

In general, there are two flavours of multiple imputation: 

1. We can either model the joint distribution of the data by means of **joint modeling** 
2. Or, we can model each variable separately by means of **fully conditional specification**

## Joint modeling
With JM, imputations are drawn from an assumed joint multivariate distribution. 

- Often a multivariate normal model is used for both continuous and categorical data, 
- Other joint models have been proposed (see e.g. Olkin and Tate, 1961; Van Buuren and van Rijckevorsel, 1992; Schafer, 1997; Van Ginkel et al., 2007; Goldstein et al., 2009; Chen et al., 2011). 

Joint modeling imputations generated under the normal model are usually robust to misspecification of the imputation model (Schafer, 1997; Demirtas et al., 2008), **although transformation towards normality is generally beneficial.**

### Procedure
1. Specify the joint model $P(Y,X)$
2. Derive $P(Y_\mathrm{mis}|Y_\mathrm{obs},X)$
3. Draw imputations $\dot Y^\mathrm{mis}$ with a Gibbs sampler

## Joint modeling
**PRO**

- The conditionals are compatible
- The statistical inference is correct under the assumed joint model
- Efficient parametrization is possible
- The theoretical properties are known

**CON**

- Having to specify a joint model impacts flexibility
- The JM can assume more than the complete data problem
- It can lead to unrealistically large models
- The assumed model may not be very close to the data


## FCS
Multiple imputation by means of FCS does not start from an explicit multivariate model. 

<div class="notepaper">
  <figure class="quote">
    <blockquote class="curly-quotes" cite="https://www.youtube.com/watch?v=qYLrc9hy0t0">
    <font color="black">
   With FCS, multivariate missing data is imputed by univariately specifying an imputation model for each incomplete variable, conditional on a set of other (possibly incomplete) variables. 
    </font>
    </blockquote>
  </figure>
</div>

- the multivariate distribution for the data is thereby implicitly specified through the univariate conditional densities 
- imputations are obtained by iterating over the conditionally specified imputation models.

### Procedure

- Specify $P(Y^\mathrm{mis} | Y^\mathrm{obs}, X)$
- Draw imputations $\dot Y^\mathrm{mis}$ with Gibbs sampler

## FCS
The general idea of using conditionally specified models to deal with missing data has been discussed and applied by many authors 

  - see e.g. Kennickell, 1991; Raghunathan and Siscovick, 1996; Oudshoorn et al., 1999; Brand, 1999; Van Buuren et al., 1999; Van Buuren and Oudshoorn, 2000; Raghunathan et al., 2001; Faris et al., 2002; Van Buuren et al., 2006. 

Comparisons between JM and FCS have been made that indicate that FCS is a useful and flexible alternative to JM when the joint distribution of the data is not easily specified (Van Buuren, 2007) and that similar results may be expected from both imputation approaches (Lee and Carlin, 2010).

### FCS in `mice`

- Specify the imputation models $P(Y_j^\mathrm{mis} | Y_j^\mathrm{obs}, Y_{-j}, X)$

  - where Y_{−j} is the set of incomplete variables except Y_j
  
- Fill in starting values for the missing data
- And iterate

## Why I prefer FCS

**PRO**

- FCS is very flexible 
- modeling remains close to the data
- one may use a subset of predictors for each column
- work very well in practice
- straightforward to explain to applied researchers

**CON**

- its theoretical properties are only known in special cases
- potential incompatibility of the collection of conditionals with the joint
- no computational shortcuts

Bold statement:

$$\text{Merging JM and FCS would be better}$$

# Merge JM and FCS

## Hybrids of JM and FCS
We can combine the flexibility of FCS with the appealing theoretical properties of JM

In order to do so, we need to partition the variables into **blocks**

- For example, we might partition $b$ blocks $h = 1,\dots,b$ as follows

  - a single block with $b=1$ would hold a **joint model**:
$$\{Y_1, Y_2, Y_3, Y_4\}, X$$
  - a quadruppel block with $b=4$ would be the `mice` algorithm
  $$\{Y_1\},\{Y_2\},\{Y_3\},\{Y_4\}, X$$

  - anything in between would be a hybrid between the joint model and the `mice` model. For example,
  $$\{Y_1, Y_2, Y_3\},\{Y_4\}, X$$  

## Why is this useful
It is not rare that sets of variable follow a deterministic relation or system. Examples are

- **Imputing squares/nonlinear effects**: In the model $y=\alpha + \beta_1X+\beta_2X^2 + \epsilon$, $X$ and $X^2$ should be imputed jointly (Von Hippel, 2009, Seaman, Bartlett & White, 2012, Vink & Van Buuren, 2013, Bartlett et al., 2015)
- **Compositional data**: Predictive ratio matching (Vink, 2015, Ch5)

$$
\begin{array}{lllllllllllll}
x_0 &=	&x_1		&+	&x_2		&+		&x_3		&+& x_4	& 		& 	& 	&\\
       &  	&= 		&   	&      		& 		&  		&& =		& 		& 	&	&\\
       &  	&x_9    	&   	&      		&		& 		&& x_5	& 		& 	&	&\\
       &  	&+		&   	&      		&		& 		&& +		& 		& 	&	&\\
       &  	&x_{10} 	&   	&      		&		& 	 	&&x_6		&= 		&x_7 	&+&x_8
\end{array}
$$

- **Multivariate PMM**: Imputing a combination of outcomes optimally based on a linear combination of covariates (Cai, Vink & Van Buuren - working paper). 

## JM embedded within FCS

b  | h | target            | predictors| type
---|---|-------------------|-----------|------
2  | 1 | $\{Y_1, Y_2, Y_3\}$ | $Y_4, X$ | mult
2  | 2 | $Y_4$ | $Y_1, Y_2, Y_3, X$     | univ

## FCS embedded within FCS 

b  | h | j | target| predictors      | type
---|---|---|-------|------------------|--------
2  | 1 | 1 | $Y_1$ | $Y_2, Y_3, Y_4, X$ | univ
2  | 1 | 2 | $Y_2$ | $Y_1, Y_3, Y_4, X$ | univ
2  | 1 | 3 | $Y_3$ | $Y_1, Y_2, Y_4, X$ | univ
2  | 2 | 1 | $Y_4$ | $Y_1, Y_2, Y_3, X$ | univ

## Benefits of blocks in `mice()`

1. Looping over $b$ blocks instead of looping over $p$ columns. 
2. Only specify $b \times p$ predictor relations and not $p^2$. 
3. Only specify $b$ univariate imputation methods instead  of $p$ methods. 
4. Ability for imputing more than one column at once
5. Simplified overall model specification
  - e.g. sets of items in scales, matching items in longitudinal data, joining data sets, etc.

## `predictorMatrix` simplification: 

Under the conventional FCS predictor specification, we could hypothesize the following `predictorMatrix`. 

```{r echo=FALSE}
nam <- c("age", "item1", "item2", "sum_items", "time1", "time2", "time3", "mean_time")
a <- matrix(c(0, 0, 0, 1, 0, 0, 0, 1,
              1, 0, 1, 0, 0, 0, 0, 1,
              1, 1, 0, 0, 0, 0, 0, 1,
              0, 1, 1, 0, 0, 0, 0, 0,
              1, 0, 0, 1, 0, 1, 1, 0,
              1, 0, 0, 1, 1, 0, 1, 0,
              1, 0, 0, 1, 1, 1, 0, 0,
              0, 0, 0, 0, 1, 1, 1, 0),
              byrow = TRUE, nrow = 8, 
            dimnames = list(nam, nam))
print(a)
```

## `predictorMatrix` simplification: 
Under the new `blocked` approach, however, we could simplify these specifications into the following blocks and predictor relations. 

```{r}
blocks <- list(age = "age", 
               A = c("item1", "item2", "sum_items"), 
               B = c("time1", "time2", "time3", "mean_time"))
```
```{r echo = FALSE}
namr <- c("age", "Items", "Time")
b <- matrix(c(0, 0, 0, 1, 0, 0, 0, 1,
              1, 0, 0, 0, 0, 0, 0, 1,
              1, 0, 0, 1, 0, 0, 0, 0),
              byrow = TRUE, nrow = 3, 
            dimnames = list(namr, nam))
print(b)
```

## An example: `brandsma`
The `brandsma` dataset (Snijders and Bosker, 2012) contains data from 4106 pupils in 216 schools. 
```{r}
d <- brandsma %>% select(sch, lpo, iqv, den)
head(d)
```

The scientific interest is to create a model for predicting the outcome `lpo` from the level-1 predictor `iqv` and the measured level-2 predictor `den` (which takes values 1-4). For pupil $i$ in school $c$ in composition notation:

$$lpo_{ic} = \beta_0 + \beta_1\mathrm{iqv}_{ic} + \beta_2\mathrm{den}_c + \upsilon_{0c}+ \epsilon_{ic}$$
where $\epsilon_{ic} \sim \mathcal{N}(0, \sigma_\epsilon^2)$ and $\upsilon_{0c} = \mathcal{N}(0, \sigma_\upsilon^2)$

## Normally in `mice`
```{r}
meth <- make.method(d)
meth[c("lpo", "iqv", "den")] <- c("2l.pmm", "2l.pmm", "2lonly.pmm")
meth
pred <- make.predictorMatrix(d)
pred["lpo", ] <- c(-2, 0, 3, 1) # -2 denotes cluster identifier
pred["iqv", ] <- c(-2, 3, 0, 1) # 3 denotes including the cluster mean for a covariate
pred["den", ] <- c(-2, 1, 1, 0) # 1 denotes fixed effects predictor
pred
```
```{r eval = FALSE}
imp <- mice(d, pred = pred, meth = meth, seed = 123, m = 10, print = FALSE)
```

## With `blocks`
We use the function `mitml::jomoImpute` and call it from within `mice`
```{r cache = TRUE}
d$den <- as.factor(d$den)
block <- make.blocks(d, "collect") # assign all vars to a single block
formula <- list(collect = list(lpo + iqv ~ 1 + (1 | sch),
                               den ~ 1))

```
We parse the `block` and `formula` objects to their respective arguments in the `mice` function
```{r eval = FALSE}
imp <- mice(d, meth = "jomoImpute", blocks = block,
            form = formula, print = FALSE, seed = 1,
            maxit = 2, m = 10, n.burn = 100)
```


## Conclusion

Imputing column in blocks is a straigtforward extension to the FCS algorithm

- Fully implemented in `mice` in `R`
- Adding blocks allows for more flexibility in cases where multivariate imputation would lead to to better inference
- Using blocks allows the modeler to remain closer to the observed data
- It is simple to specify hybrid models

**To do**

- We're still working on the documentation that details a variety of use cases

**Potentially interesting**

Integrate the hybrid (JM/FCS) approach with the blocked sequential regression multivariate imputation approach by Zhu (2016), which in turn generalizes the monotone block approach of Li et al. (2014). 


## References

The multilevel example is detailed in [Van Buuren (2018), Chapter 7.10.4](https://stefvanbuuren.name/fimd/sec-mlguidelines.html)

See [www.gerkovink.com/London2019/](https://www.gerkovink.com/London2019) for an detailed overview of all references. 